---
title: "Pre-Trained Annotation Models for TALL"
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 150
)
```

## Overview

**TALL - *Text Analysis for ALL*** (https://github.com/massimoaria/tall) is an R Shiny application that provides a comprehensive suite of methodologies for various text analysis tasks. It is designed to meet the needs of researchers who may lack extensive programming skills, offering a versatile and user-friendly tool for analyzing textual data. With TALL, users can apply a wide range of text analysis techniques, making it easier to extract valuable insights from textual data in an efficient and accessible manner.

**TALL** leverages the `udpipe` library for preprocessing, tokenization, lemmatization, and parsing of raw texts. The `udpipe` R package allows users to easily tokenize, tag, lemmatize, and perform dependency parsing on texts in multiple languages by providing convenient access to pre-trained annotation models. Built-in udpipe models are based on Universal Dependencies (UD) version 2.5 and had not been updated in some time.

This repository, `udpipe.models`, provides updated pre-trained NLP language models, trained on gold standard annotated corpora from the [Universal Dependencies](https://universaldependencies.org/) (UD) project, version 2.15, using the `udpipe` R package ([CRAN link](https://CRAN.R-project.org/package=udpipe)).

The models are intended for use in Natural Language Processing (NLP) tasks, including tokenization, parsing, and tagging. This repository serves as a resource for researchers and developers working with TALL ([Tall GitHub repository](https://github.com/massimoaria/tall)).

## Purpose

The primary objectives of this repository are:

- To provide updated pre-trained `udpipe` models for a wide range of languages, aligned with the latest version of the UD corpora (2.15).
- To facilitate reproducible NLP research and applications by making these models readily available.
- To support projects that use Universal Dependencies data for tokenization, parsing, and tagging tasks in R and other environments.

## Data Source

The gold standard annotated corpora used to train these models are sourced from the Universal Dependencies project ([Universal Dependencies GitHub repository](https://github.com/UniversalDependencies)).

Universal Dependencies is a community-driven initiative that offers high-quality annotated corpora in the standardized CoNLL-U format for many languages. The models in this repository are trained specifically on data from version 2.15.

## Available Pre-Trained Models

```{r, echo=FALSE, results="asis"}
# Carichiamo i pacchetti necessari
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(knitr))

load("data/available_models.rdata")
df %>% 
  mutate(file = paste0(file,"-ud-2.15.udpipe"),
  language_name = tools::toTitleCase(gsub("_"," ",language_name))) %>%
  kable("markdown", col.names = c("Language", "Treebank", "Contributors", "Description", "File", "Link")) 

```


## Model Accuracy

We present the tokenizer, tagger, lemmatizer and parser performance, measured on the testing portion of the data, evaluated using the gold tokenization. The results are F1 scores

```{r, echo=FALSE, results="asis"}
load("data/model_accuracy.rdata")

row.names(accuracy) <- NULL
accuracy %>% 
  select(-file) %>%
  kable("markdown")
```


## Acknowledgements

The models were trained using [UDpipe R package](https://CRAN.R-project.org/package=udpipe) on [Universal Dependencies 2.15 treebanks](https://universaldependencies.org/).


## Structure

The repository is organized as follows:

- **`2.15/`**: Contains pre-trained `udpipe` models for individual languages. Each model file is named according to the corresponding language and the version of the Universal Dependencies data used.
- **`data/`**: Contains rdata files to build up tables in README.
- **`README.rmd`**: This document provides an overview of the repository's contents and purpose (rmarkdown format).
- **`README.md`**:  This document provides an overview of the repository's contents and purpose (github format).
- **`README.html`**:  This document provides an overview of the repository's contents and purpose (html format).

## Using the Models

To use these models, simply install **TALL**, which will automatically handle the downloading of the pre-trained model required for analyzing your corpus.

```r
# Install the remotes package if not already installed (uncomment the next line removing #)
# install.packages("remotes")

remotes::install_github("massimoaria/tall")

library(tall)

tall()
```